NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-01-17 main.tex
Update text

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. the importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access is still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. The Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best models for specific applications and accelerates the development of simulation parameters and methodology. The programmatic access enables users to apply data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a wide range of fields covering academia and industry, from cell membrane biology and lipid nanoparticle formulations to computational chemistry and machine learning. As two usage examples of what is already possible, we demonstrate here (i) how data-driven analysis of water anisotropic diffusion in all the membrane systems available in the Databank can extend the scope of MD simulations to magnetic resonance imaging (MRI) and pharmacokinetics, and (ii) how the Databank allows its users to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Wider adaptation of the NMRlipids Databank will open even more possibities.

@@

\subsection{Accessing NMRlipids databank}
The content of the NMRlipids databank is easily searchable and browsable via graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}), which can be also used to visualize the systems, compare basic properties between diffent systems and experiments, and rank simulations based on their quality evaluated against experimental data.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=88mm]{Figures/flowchart2.pdf}
    \caption{ A) Flowchart for analysing properties from all MD simulations in the NMRlipids databank.
    B) Flowchart for accessing the results analyzed from the NMRlipids databank.
    C) Area per lipids of POPC and POPE lipid bilayers predicted by different force fields.  
    }
    \label{fig:POPCPOPEapls}
\end{figure}

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding of how closely simulations resemble the reality is crucial when assessing the reliability and significance of MD simulation results, as well as when using atomistic simulations in parametrization of coarse grained models. However, universal quality measures for MD simulations are lacking, and a prevailing force field that would perform best in all situations has not been found for membranes. For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. 
%and the complicated landscape of force field quality for membranes hamper the applicability of such simulations, resulting in substantial controversies~\cite{antila22b}. 
%For example, previous quality evaluations of membrane simulations have concluded that 
On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. Furthermore, Fig.~\ref{fig:POPCPOPEapls}B illustrates divergent predictions for the difference in packing between POPC and POPE membranes from different force fields. To streamline the selection of best simulations for specific applications, and to guide the development of atomistic and coarse grained parameters, quantitative quality measures for lipid bilayer simulations are defined in the NMRlipids databank. Together with the API, these will enable rapid and automatic quality evaluation of simulations in the NMRlipids databank, thereby promoting the selection of best simulations for specific applications without laborious force field evaluation, and automatic parametrization procedures for atomistic and coarse grained simulations.
%that can be used to rapidly find the best available parameters .

@@

Ranking based on C--H bond order parameter quality for the best performing simulations is shown in Fig.~\ref{fig:quality}A, and for the best performing POPE lipid bilayer simulations in Fig.~\ref{fig:quality}B. Direct comparisons with experiments are shown for selected simulations in Figs.~\ref{fig:quality}C--E. 
%The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls} showing area per lipids from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. 
This can be used to select the the simulations that best caputer area per lipids for POPE and POPC lipids in Fig.~\ref{fig:POPCPOPEapls}B because the area per lipid correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F).
%, simulations with the best predictions for area per lipids can be selected based on order parameter quality evaluation. 
Quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} reveal that Slipids ranks 1\textsuperscript{st} for POPC with a clear difference to others, and 2\textsuperscript{nd} for POPE with only marginally lower quality than GROMOS-CKP. Simulations with CHARMM36 and GROMOS-CKP predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, Slipids parameters is the best available choice for studies where packing effects of PE lipids are relevant

@@

\subsection{Detecting rare phenomena using NMRlipids databank: Cholesterol flip--flops}
Lipid flip--flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip--flops are slow when not facilitated by proteins, occurring spontaneously on the timescale of hours or days, while cholesterol, diacylglycerol and ceramide flip-flops are faster. Still, the reported timescales range from minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, yet atomistic simulations reporting cholesterol flip--flop events have been published recently~\cite{gu19,javanainen19,baral20}. 
%However, systematic studies on how cholesterol flip-flop rate depends on membrane properties are still beyond the current computational resources available for a single research group. 
These studies report an increase in cholesterol flip--flop rates with increasing acyl chain unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but correlations between cholesterol flip--flop rates and membrane properties have not been systematically studied. Here, we demonstrate that the large amount of simulation data available in the NMRlipids databank, along with programmatic access providing the option to flexibly implement varied queries of the data, enables anybody to perform such analyses.

Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol) in the NMRlipids databank. Cholesterol flip--flop rates range between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$. These values are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9\,s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Thus, available data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we use the wide range of available simulations to analyse how cholesterol flip--flop rate depends on membrane properties. Cholesterol flip--flop rates and their histograms are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B--D. The results reveal a non-linear correlation between cholesterol flip--flop rate and membrane packing. Flip--flop rates increase by an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip--flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip--flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

\subsection{Extending the scope of MD simulations to new fields using the NMRlipids databank: Water diffusion anisotropy in membrane systems}
The anisotropy in water diffusion in parallel and perpendicular directions with respect to membranes plays a role in the drug translocation through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21}, and in MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze anistropic diffusion of water since only few membrane permeation events for water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of sufficient amount of data challenging or impossible. Here, we show that the NMRlipids databank can be used to systemically analyze how the anisotropic water diffusion in multilamellar membrane systems depends on membrane properties.

@@

To this end, we calculated the water permeability through membranes from all simulations in the NMRlipids databank. The resulting non-zero values range between 0.3 and 322\,\textmu{}m/s with the mean and median of 14\,\textmu{}m/s and 8\,\textmu{}m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022} and have the same order of magnitude as experimentally determined diffusive permeability coefficients, but are on average larger than the values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Observed permeabilities and their histogrammed values are shown in Figs.~\ref{fig:permeability} B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability} B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers also depends on membrane properties also according previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%, although weak decrease for the latter two may be visible.
%
%
%
%%Several models on how water dynamics depends on membrane properties have been proposed based on different experimental and %%theoretical methods, yet consensus 
%%%on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, 
%%has not been reached~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI}. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\mathrm{\perp}/D_{||}$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\mathrm{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. 
%Large anisotropy values are understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. 
Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\mathrm{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

In summary, our results suggest that the bilayer packing has a substantial effect on anisotropic water diffusion in multi-membrane lipid systems. Several folds larger anisotropy in membranes with higher lateral density is expected to play a role in pharmacokinetic models not only for water but also for other hydrophilic molecules~\cite{nitsche19}. Furthermore, the enhanced understanding of this anisotropy may help in developing new (functional) MRI imaging methods~\cite{topgaard20}.

@@

\subsection{Calculation of X-ray scattering form factors}
X-ray scattering form factors were calculated with standard equation for symmetric lipid bilayers~\cite{ollila16}
\begin{equation}\label{FFsimpl}
F(q)=\int_{-D/2}^{D/2}\Delta \rho_e(z) \cos(zq_z) {\rm d}z,
\end{equation}
where $\Delta \rho_e(z)$ is the difference between total and solvent electron densities, and $D$ is the simulation box size in the $z$-direction (normal to the membrane). For the calculation of density profiles, atom coordinates were first centred around the centre of mass of lipid molecules for every time frame, and a histogram of these centred positions weighted with the number of electrons in each atom was then calculated with the bin width of $1/3$~\AA{}. The script to calculate form factors for all simulations in the NMRlipids databank is available at \texttt{Scripts/AnalyzeDatabank/calc\_FormFactors.py}. The resulting form factors are stored for all simulations in files named \texttt{FormFactor.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

\subsection{Calculation of a bilayer area per lipid and thickness}
Area per lipids of bilayers were calculated by dividing the time-averaged area of the simulation box with the total number of lipid molecules in the simulation. The script that calculates area per lipids from all simulations in the NMRlipids databank repository is available at \texttt{Scripts/AnalyzeDatabank/calcAPL.py} in the NMRlipids databank repository. The resulting area per lipids are stored for all simulations in files named \texttt{apl.json} at folders in \texttt{/Data/Simulations}.

@@

>>>>

In contrast to experimental structural biology, where standard protocols to share and quality-evaluate the resolved structures are firmly established \textit{via} the Protein Data Bank (PDB)~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\url{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access is still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. Recent advances using machine learning approaches utilizing publicly available databanks, for example to solve protein structures~\cite{jumper21}, emphasize the increasing importance of such resources. 
%The profound, and sometimes unexpected, effect that such open data sharing and quality evaluation can have on the field cannot be understated. At the very least, open access data will reduce the man and computational hours originating from double work, reducing the environmental impact. Beyond that, by providing the training data, the PDB laid the foundation for the machine learning algorithms now closing into solving the protein folding problem and the strides those machine learning tools have facilitated in the biomedical sciences since.

Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. The programmatic access enables users to apply data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a wide range of fields covering academia and industry, from cell membrane biology and lipid nanoparticle formulations to computational chemistry and machine learning. As two usage examples of what is already possible, we demonstrate here (i) how data-driven analysis of water anisotropic diffusion in all the membrane systems available in the Databank can extend the scope of MD simulations to magnetic resonance imaging (MRI) and pharmacokinetics, and (ii) how the Databank allows its users to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Wider adaptation of the NMRlipids Databank will open even more possibities. Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best models for specific applications and accelerates the development of simulation parameters and methodology.

@@

\subsection{Accessing NMRlipids databank}\label{section:access}
The content of the NMRlipids databank is easily searchable and browsable via graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}), which can be also used to visualize the systems, compare basic properties between diffent systems and experiments, and rank simulations based on their quality evaluated against experimental data.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=88mm]{Figures/flowchart2.pdf}
    \caption{ A) Flowchart for analysing properties from all MD simulations in the NMRlipids databank.
    B) Flowchart for accessing the results analyzed from the NMRlipids databank.
    C) Area per lipids of POPC and POPE lipid bilayers predicted by different force fields at 310\,K. The points from the best performing simulations in the NMRlipids databank quality evaluation for each lipid (Figs.~\ref{fig:quality} and~\ref{fig:POPC_POPE_dataSI}) are surrounded by black circles. 
    }
    \label{fig:POPCPOPEapls}
\end{figure}

\subsection{Selection of best simulation parameters for applications using the NMRlipids databank}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
Understanding of how closely simulations resemble the reality is crucial when assessing the reliability and significance of MD simulation results, as well as when using atomistic simulations in parametrization of coarse grained models. However, universal quality measures for MD simulations are lacking, and a prevailing force field that would perform best in all situations has not been found for membranes. For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. Furthermore, Fig.~\ref{fig:POPCPOPEapls}B illustrates divergent predictions for the difference in packing between POPC and POPE membranes from different force fields. To streamline the selection of best simulations for specific applications, and to guide the development of atomistic and coarse grained parameters, quantitative quality measures for lipid bilayer simulations are defined in the NMRlipids databank. Together with the API, these will enable rapid and automatic quality evaluation of simulations in the NMRlipids databank, thereby promoting the selection of best simulations for specific applications without laborious force field evaluation, and automatic parametrization procedures for atomistic and coarse grained simulations.
%that can be used to rapidly find the best available parameters .

@@

Ranking based on C--H bond order parameter quality for the best performing simulations is shown in Fig.~\ref{fig:quality}A, and for the best performing POPE lipid bilayer simulations in Fig.~\ref{fig:quality}B. Direct comparisons with experiments are shown for selected simulations in Figs.~\ref{fig:quality}C--E. 
%The power of the NMRlipids databank in selecting the best simulation parameters for a specific application is demonstrated for mixtures of PE and PC lipids in Fig.~\ref{fig:POPCPOPEapls} showing area per lipids from available POPC and POPE bilayer simulations in the databank from different force fields at 310\,K. Predictions from different force fields deviate significantly in terms of absolute values and the slope of decrease upon addition of POPE. 
This can be used to select the the simulations that best capture area per lipids for POPE and POPC lipids in Fig.~\ref{fig:POPCPOPEapls}B because the area per lipid correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality} F).
%, simulations with the best predictions for area per lipids can be selected based on order parameter quality evaluation. 
Quality evaluation in Fig.~\ref{fig:quality} and direct comparison in Fig.~\ref{fig:POPC_POPE_dataSI} reveal that Slipids ranks 1\textsuperscript{st} for POPC with a clear difference to others, and 2\textsuperscript{nd} for POPE with only marginally lower quality than GROMOS-CKP. Simulations with CHARMM36 and GROMOS-CKP predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, Slipids parameters is the best available choice for studies where packing effects of PE lipids are relevant

@@

\subsection{Detecting rare phenomena using NMRlipids databank: Cholesterol flip--flops}
Lipid flip--flops from one bilayer leaflet to another play an important role in lipid trafficking and regulating membrane properties~\cite{vanmeer08}. Phospholipid flip--flops are slow when not facilitated by proteins, occurring spontaneously on the timescale of hours or days, while cholesterol, diacylglycerol and ceramide flip-flops are faster. Still, the reported timescales range from minutes to sub-millisecods~\cite{vanmeer08,steck12,parisio16,gu19}. These timescales were previously accessible only by coarse-grained simulations or free energy calculations~\cite{parisio16}, yet atomistic simulations reporting cholesterol flip--flop events have been published recently~\cite{gu19,javanainen19,baral20}. These studies report an increase in cholesterol flip--flop rates with increasing acyl chain unsaturation level and decreasing cholesterol concentration~\cite{gu19,javanainen19}, but the amount of data in these individual studies was not sufficient to systematically assess correlations between cholesterol flip--flop rates and membrane properties. Here, we demonstrate that the analysis of such correlation is now accessible to all by using the API to access the large amount of MD simulation data in the NMRlipids databank as described in section~\ref{section:access}.

Using workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A, we first calculated the cholesterol flip--flop rates from all the simulations available in the NMRlipids databank. Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol) in the NMRlipids databank. Cholesterol flip--flop rates range between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$. These values are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9$\cdot$10$^{-6}$\,\textmu{}s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Thus, available data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we use the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B and the results for flip--flop rates from the NMRlipids databank to analyse how they depend on membrane properties. Cholesterol flip--flop rates and their histograms are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B--D. The results reveal a non-linear correlation between cholesterol flip--flop rate and membrane packing. Flip--flop rates increase by an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip--flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip--flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

\subsection{Extending the scope of MD simulations to new fields using the NMRlipids databank: Water diffusion anisotropy in membrane systems}
The anisotropy in water diffusion in parallel and perpendicular directions with respect to membranes can be related to the translocation of drugs  through biological material, particularly in skin \cite{hansen13,wen18,nitsche19,roberts21}, and to the formation of signals in diffusion tensor MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze the anisotropic diffusion of water since only few membrane permeation events for water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of sufficient amount of data challenging. Here, we show that the API access to the data in NMRlipids databank enables systematic analyses on how the anisotropic diffusion of water depends on membrane properties in multilamellar membrane systems, thereby extending the applications of MD simulations to new fields.

@@

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The resulting non-zero values range between 0.3 and 322\,\textmu{}m/s with the mean and median of 14\,\textmu{}m/s and 8\,\textmu{}m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022} and have the same order of magnitude as experimentally determined diffusive permeability coefficients, but are on average larger than the values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}b, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability} B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability} B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers also depends on membrane properties also according previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%, although weak decrease for the latter two may be visible.
%
%
%
%%Several models on how water dynamics depends on membrane properties have been proposed based on different experimental and %%theoretical methods, yet consensus 
%%%on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, 
%%has not been reached~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\mathrm{\perp}/D_{||}$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\mathrm{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. 
%Large anisotropy values are understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. 
Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\mathrm{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

In summary, our results suggest that the bilayer packing has a substantial effect on anisotropic water diffusion in multi-membrane lipid systems. Several folds larger anisotropy in membranes with higher lateral density is expected to play a role in pharmacokinetic models not only for water but also for other hydrophilic molecules~\cite{nitsche19}. Furthermore, the enhanced understanding of this anisotropy may help in developing new diffusion tensor based MRI imaging methods where signals originate from anisotropic diffusion of water in biological material~\cite{topgaard20}.

@@

\subsection{Calculation of X-ray scattering form factors}
X-ray scattering form factors were calculated with the standard equation for symmetric lipid bilayers~\cite{ollila16}
\begin{equation}\label{FFsimpl}
F(q)=\int_{-D/2}^{D/2}\Delta \rho_e(z) \cos(zq_z) {\rm d}z,
\end{equation}
where $\Delta \rho_e(z)$ is the difference between total and solvent electron densities, and $D$ is the simulation box size in the $z$-direction (normal to the membrane). For the calculation of density profiles, atom coordinates were first centred around the centre of mass of lipid molecules for every time frame, and a histogram of these centred positions, weighted with the number of electrons in each atom, was then calculated with the bin width of $1/3$~\AA{}. Electron density profiles were then calculated as an average of these histograms over the time frames in simulations. The script to calculate form factors for all simulations in the NMRlipids databank is available at \texttt{Scripts/AnalyzeDatabank/calc\_FormFactors.py}. The resulting form factors are stored for all simulations in files named \texttt{FormFactor.json} at folders in \texttt{/Data/Simulations} in the NMRlipids databank repository.

\subsection{Calculation of a bilayer area per lipid and thickness}
Area per lipids of bilayers were calculated by dividing the time-averaged area of the simulation box with the total number of lipid and surfactant molecules (see the list in table~\ref{tab:abbreviations}) in the simulation. The script that calculates area per lipids from all simulations in the NMRlipids databank repository is available at \texttt{Scripts/AnalyzeDatabank/calcAPL.py} in the NMRlipids databank repository. The resulting area per lipids are stored for all simulations in files named \texttt{apl.json} at folders in \texttt{/Data/Simulations}.

@@

\subsection{PCA analysis of equilibration of simulations}
